Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-06 Manuscript/TEM_draft.tex
Results section update

\subsubsection{Summarizing the comparisons}
Finally after doing all theses %stupid
50 simulations and comparisons, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$) on the ability to recover the correct topology in a Total Evidence framework. Additionally, we also assessed the effect of the different methods (i.e. Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution).% and the joint effect of the our missing data parameters and the methods.
% TG: the following part might be useless (complete repetition of above)
 To assess the ability to recover the "correct" tree topology on our 50 simulations, we used the scores for both normalised metrics (Robinson-Foulds and Triplets distance). We looked at the following effects:
\begin{enumerate}
\item{The three missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$);}
\item{The four "methods" (Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution);}
%\item{The joint effect of the missing data parameters and the methods.} actually this is implied in both already.
\end{enumerate}
% TG: end of the repetition

@@

%Explaining the methods comparisons (ANOVA equivalent=among)
Finally, for assessing the effect of the method on the ability to recover the "correct" topology, we compared the similarity between the methods. We calculated the Bhattacharyya Coefficient between each 125 normalised metrics scores (Robinson-Foulds and Triplets distance) for the 6 pair of methods (e.g. Maximum Likelihood\textit{versus} Bayesian consensus). We then reported the minimum, the 1st quartile, the median, the mean, the 3rd quartile and the maximum of the probabilities of overlap between the methods (i.e. the Bhattacharyya Coefficients) in table ~\ref{Tab_Results-Difference_methods}. When two methods perform similarly in their ability to recover the "correct" tree topology (i.e. they display similar normalised metrics scores for each comparison between the "best" tree and the "missing data" trees), we expect the Bhattacharyya Coefficients to be close to 1. On the opposite, if the two methods differ, we expect the Bhattacharyya Coefficients to be close to 0.

@@

% NC: Split into three sections to answer three questions

@@

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/AllMethods-RF+Tr-colour.pdf}
\caption{Comparison between the effect of missing data and the tree inference method on topological recovery. The amount of missing data for each parameter is represented on the x axis. The difference in topologies is represented on the y axis, using standardized Robinson-Foulds distance (upper row) and standardized Triplets distance (lower row). Points represent the modal value of each distribution; thick solid and thin dashed lines represents the 50\% and 95\% confidence intervals or the distributions respectively. Analyses using the Maximum Likelihood trees are represented in black, the Bayesian consensus %MCC?
trees in red, the Maximum Likelihood bootstrap trees in green and the Bayesian posterior tree distribution in blue.}
\label{Fig_Results-permeth_perparam} %Differences within a subset of parameters and between methods
\end{figure}

% NC: Y axes should be standardised Robinson Foulds
% NC: X-axes should have (Mc) etc in brackets after the names
% NC: Also should probably have the X axes read : Missing character (Mc, %) and remove % from tick labels

%NC: I would rewrite these figure labels as follows. I prefer the details of colours and things to be in brackets not in the text. Rewrite the others in a similar fashion.
%NC: Remeber it's not topology, but topological recovery that you are measuring
%The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (red), Maximum Likelihood bootstrap trees (green) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter (Mc, ML, MF) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: standardized Robinson-Foulds distance (upper row) and standardized Triplets distance (lower row). The graph shows the modal value (points), the 50\% and 95\% confidence intervals (thick solid lines) and the distributions (thin dashed lines), of the tree comparison metric for each missing data parameter and tree inference method.

As the amount of missing data in the morphological part of the total evidence matrix increases, our ability to recover the correct topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{C}$ or $M_{F}$), the method of tree inference (Maximum Likelihood or Bayesian) or the tree comparison metric used (standardized Robinson Foulds or standardized Triplets distance). 
% NC: Results are for YOUR results. compare it to others in the Discussion only
% NC: I've rephrased below as it was getting confused between what is GOOD and BAD for topological recovery and the difference between the parameters and the methods. I've tried to clarify.
% NC: Note that you need to be clear what the methods are outperforming...
However, the different missing data parameters and tree inference methods do not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam}). % And figure 3 too right?

@@

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/ML+Baycon-AllParam-RF+Tr-BW.pdf}
\caption{Trend of the effect of missing data on topological recovery on the Maximum Likelihood and the Bayesian consensus trees. The amount of missing data per parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is represented along the x axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data for each parameter. The topology is represented on the y axis, both using Robinson-Foulds distance (upper row) and Triplets distance (lower row). Points represent the modal value of each distribution ; thick solid and thin dashed lines represents respectively the 50\% and 95\% confidence intervals or the distributions. The Maximum Likelihood trees are represented in black and the Bayesian consensus trees in grey.}
\label{Fig_Results-global_perparam} %Differences between all the parameters and between two methods (ML vs BC)
\end{figure}

@@

\subsection{Effects of the missing data parameters on topological recovery}

%Hide this figure for 'fast' build
%\begin{figure} 
%\centering
%    \includegraphics[width=1\textwidth]{Figures/In_main/PairwiseComp-Baycon-RF+Tr-colour.pdf}
%\caption{Pairwise Bhattacharyya Coefficients within the Bayesian consensus trees. The pairwise trees comparisons are represent on both axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data for each parameter. The matrix represents the values of pairwise Bhattacharyya Coefficients going from green (1) to red (0). A. Results for the Normalised Robinson-Foulds distance. B. Results with the for Normalised Triplets distance.}
\label{Fig_Results-paircomp_within}
%\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

@@

\begin{table}[ht]
\caption{Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised distance between the "best" tree and the "missing data".}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.00 & 0.00 & 0.10 & 0.20 & 0.32 & 1.00 \\ 
    Bayesian consensus              & $Tr^2$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.03 & 0.54 & 0.69 & 0.64 & 0.77 & 0.98 \\ 
    Bootstraps                      & $Tr^2$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.02 & 0.74 & 0.80 & 0.79 & 0.89 & 0.98 \\ 
    Bayesian posterior trees        & $Tr^2$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} & $RF^1$ & 0.00 & 0.00 & \textbf{0.00} & \textbf{0.01} & 0.01 & 0.04 \\ 
    bootstraps                      & $Tr^2$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} & $RF^1$ & 0.00 & 0.00 & \textbf{0.01} & \textbf{0.02} & 0.04 & 0.11 \\ 
    Bayesian posterior trees        & $Tr^2$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Boostraps \textit{vs.}          & $RF^1$ & 0.50 & 0.77 & \textbf{0.85} & \textbf{0.85} & 0.96 & 1.00 \\ 
    Bayesian posterior tree         & $Tr^2$ & 0.91 & 0.96 & \textbf{0.98} & \textbf{0.97} & 0.99 & 1.00 \\ 
   \hline
\end{tabular}
   $^1RF$=Normalised Robison-Fould distance ; $^2Tr$=Normalised Triplets distance
\label{Tab_Results-Difference_methods}
\end{table}

@@

% NC: Make sure you're consistent in how you refer to things ie Maximum Likelihood bootstrap trees, not Bootstraps
%---------------------------------------------
%
%       DISCUSSION
%
%---------------------------------------------

>>>>

\subsubsection{Summarizing the comparisons}
Finally after doing all theses %stupid
50 simulations and comparisons, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$) on the ability to recover the correct topology in a Total Evidence framework. Additionally, we also assessed the effect of the different methods (i.e. Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution).% and the joint effect of the our missing data parameters and the methods.
% TG: the following part might be useless (complete repetition of above)
 To assess the ability to recover the "correct" tree topology on our 50 simulations, we used the scores for both normalised metrics (Robinson-Foulds and Triplets distance). We looked at the following effects:
\begin{enumerate}
\item{The three missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$);}
\item{The four "methods" (Maximum Likelihood, Bayesian consensus, Bootstraps and Bayesian posterior trees distribution);}
%\item{The joint effect of the missing data parameters and the methods.} % actually this is implied in both already.
\end{enumerate}
% TG: end of the repetition

@@

%Explaining the methods comparisons (ANOVA equivalent=among)
Finally, for assessing the effect of the method on the ability to recover the "correct" topology, we compared the similarity between the methods. We calculated the Bhattacharyya Coefficient between each 125 normalised metrics scores (Robinson-Foulds and Triplets distance) for the 6 pair of methods (e.g. Maximum Likelihood\textit{versus} Bayesian consensus). We then reported the minimum, the 1st quartile, the median, the mean, the 3rd quartile and the maximum of the probabilities of overlap between the methods (i.e. the Bhattacharyya Coefficients) in table~\ref{Tab_Results-Difference_methods}. When two methods perform similarly in their ability to recover the "correct" tree topology (i.e. they display similar normalised metrics scores for each comparison between the "best" tree and the "missing data" trees), we expect the Bhattacharyya Coefficients to be close to 1. On the opposite, if the two methods differ, we expect the Bhattacharyya Coefficients to be close to 0.

@@

% NC: Split into three sections to answer three questions
% TG: Ok I removed all the figures + table on top of the section in here and all the text down. We might to have a think of where to anchor the figures later on.

@@

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/AllMethods-RF+Tr-colour.pdf}
\caption{The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (red), Maximum Likelihood bootstrap trees (green) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method. }
\label{Fig_Results-permeth_perparam} %Differences within a subset of parameters and between methods
\end{figure}

% NC: Y axes should be standardised Robinson Foulds
% NC: X-axes should have (Mc) etc in brackets after the names
% NC: Also should probably have the X axes read : Missing character (Mc, %) and remove % from tick labels
% TG: Ok, will fix later on since it's all scripted (no illustrator cheats)

%NC: I would rewrite these figure labels as follows. I prefer the details of colours and things to be in brackets not in the text. Rewrite the others in a similar fashion.
%NC: Remeber it's not topology, but topological recovery that you are measuring
%The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (red), Maximum Likelihood bootstrap trees (green) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/ML+Baycon-AllParam-RF+Tr-BW.pdf}
\caption{Effect of increasing missing data on topological recovery using Maximum Likelihood trees (black) and Bayesian consensus trees (grey). The percentage of missing data (gradient colour from 0\% (white), 10\%, 25\%, 50\% and 75\% (black)) for each parameter ($M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$(lower line)) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.} 
\label{Fig_Results-global_perparam} %Differences between all the parameters and between two methods (ML vs BC)
\end{figure}

%Hide this figure for 'fast' build
%\begin{figure} 
%\centering
%    \includegraphics[width=1\textwidth]{Figures/In_main/PairwiseComp-Baycon-RF+Tr-colour.pdf}
%\caption{Effect of missing data on topological recovery using Bayesian consensus trees. The percentage of missing data (gradient colour from 0\% (white), 10\%, 25\%, 50\% and 75\% (black)) for each parameter ($M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$(lower line)) is shown on both x and y axis. Topological recovery is represented by the probability of the two different tree comparison metrics (Normalised Robinson-Foulds distance (A - right) and Normalised Triplets distance (B - left)) distribution overlap with the "best" tree using the the Bhattacharyya Coefficients (low probability of overlap - red ; high probability of overlap - green).}
\label{Fig_Results-paircomp_within}
%\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

\begin{table}[ht]
\caption{Bhattacharyya Coefficients of the pairwise method comparisons, each of which corresponds to the normalised distance between the "best" tree and the "missing data".}
\centering
\begin{tabular}{lccccccc}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.00 & 0.00 & 0.10 & 0.20 & 0.32 & 1.00 \\ 
    Bayesian consensus              & $Tr^2$ & 0.34 & 0.49 & 0.61 & 0.62 & 0.75 & 1.00 \\ 
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.03 & 0.54 & 0.69 & 0.64 & 0.77 & 0.98 \\ 
    Maximum Likelihood bootstraps   & $Tr^2$ & 0.08 & 0.57 & 0.65 & 0.64 & 0.73 & 0.82 \\ 
    Maximum Likelihood \textit{vs.} & $RF^1$ & 0.02 & 0.74 & 0.80 & 0.79 & 0.89 & 0.98 \\ 
    Bayesian posterior trees        & $Tr^2$ & 0.21 & 0.67 & 0.73 & 0.72 & 0.77 & 0.84 \\ 
    Bayesian consensus \textit{vs.} & $RF^1$ & 0.00 & 0.00 & \textbf{0.00} & \textbf{0.01} & 0.01 & 0.04 \\ 
    Maximum Likelihood bootstraps   & $Tr^2$ & 0.08 & 0.38 & 0.59 & 0.57 & 0.73 & 0.84 \\ 
    Bayesian consensus \textit{vs.} & $RF^1$ & 0.00 & 0.00 & \textbf{0.01} & \textbf{0.02} & 0.04 & 0.11 \\ 
    Bayesian posterior trees        & $Tr^2$ & 0.21 & 0.36 & 0.56 & 0.55 & 0.74 & 0.87 \\ 
    Bayesian posterior tree \textit{vs.}  & $RF^1$ & 0.50 & 0.77 & \textbf{0.85} & \textbf{0.85} & 0.96 & 1.00 \\ 
    Maximum Likelihood bootstraps   & $Tr^2$ & 0.91 & 0.96 & \textbf{0.98} & \textbf{0.97} & 0.99 & 1.00 \\ 
   \hline
\end{tabular}
   $^1RF$=Normalised Robison-Fould distance ; $^2Tr$=Normalised Triplets distance
\label{Tab_Results-Difference_methods}
\end{table}

%Results plan:
% A-Change in topology
%   A.1-Missing data reduce topological recovery regardless the method, etc... -> figure topological recovery ~ parameters(all - no interactions) + methods(all) + metrics(all)
%   A.2-Within a particular method, missing data reduce topological recovery   -> figure topological recovery ~ parameters(all + interaction) + methods(ML/baycon) + metrics(all)
% B-"Significance" (i.e. Bhattacharrya Coefficients as a two-sided t-test. Or in other words, does the distributions between two stuff (within/among) differ, regardeless who's better or not:
%   B.1-Effect of the parameters  -> figure matrix
%   B.2-Effect if the method      -> table matrix

% A-Change in topology
% A.1
As the amount of missing data in the morphological part of the total evidence matrix increases, our ability to recover the correct topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $M_{C}$), the method of tree inference (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets distance). 
% NC: Results are for YOUR results. compare it to others in the Discussion only
% NC: I've rephrased below as it was getting confused between what is GOOD and BAD for topological recovery and the difference between the parameters and the methods. I've tried to clarify.
% NC: Note that you need to be clear what the methods are outperforming...
However, the different missing data parameters and tree inference methods do not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). % And figure 3 too right? TG: Yep

% A.2
The Normalised Robinson-Foulds distance calculate for the trees inferred in Maximum Likelihood framework decreases with the increase of missing data in our three parameters ($M_{L}$, $M_{F}$ or $M_{C}$) with a modal Normalised Robinson-Fould score always below 0.5 when 75\% of the living taxa have no data ($M_{L}$=75\% - Fig.~\ref{Fig_Results-global_perparam}). However, in Bayesian inference, the ability of the method (the Bayesian consensus tree) to recover a "correct" topology plateaus at a minimal Normalised Robinson-Fould score of 0.69 (Fig.~\ref{Fig_Results-global_perparam} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} for the values distribution). The results of the topological recovery with the parameters combination in both Normalised Robinson-Foulds and Triplets distances for the Maximum Likelihood bootstrap trees and the Bayesian posterior distribution trees are available in the \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}.

% B-"Significance"
% B.1 - within a method
%All parameters combinations are different than the "best" tree (always red) in both metrics.
The lowest probability of overlap (low Bhattacharrya Coefficient values) between pairwise distributions of parameters combinations ($M_{L}$, $M_{F}$ and $M_{C}$ with 0\%, 10\%, 25\%, 50\% or 75\% missing data) is always between the comparison of Normalised distance (Robinson-Foulds or Triplets) of the "missing data" tree with no missing data ($M_{L}$=0\%, $M_{F}$=0\% and $M_{C}$=0\%) and all the other "missing data" trees.
%The parameters combinations are less dissimilar in Triplets than in RF (less red).
However, pairwise comparisons are more dissimilar (low Bhattacharrya Coefficient values) in the Normalised Robinson-Fould distance results than in the Normalised Triplets distance results.
%The ML parameter (0% and 10%) are more different in RF and Triplets (red stripes at the base). The ML parameter (50, 75) are the more similar (green upper part) in RF.
The Normalised Robinson-Fould distance for the comparisons between the trees with a low amount of missing data for the living taxa ($M_{L}$=0\% and 10\%) shows the most dissimilarities with the other parameter combinations.
%The combination of the ML parameter (0% and 10%) and the MC parameter (0, 10, 25) are the most different (details on the red stripes).

%Check supplementary for the rest
The results of the probability of overlap between the pairwise distributions of Normalised Robinson-Foulds and Triplets distances of the Maximum Likelihood trees, the Maximum Likelihood Bootstrap trees and the Bayesian posterior trees distribution is available in the \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}.

% B.2 - among methods
%The Results for the Bayesian consensus tree in RF are the more dissimilar to the results of the Maximum Likelihood bootstrap trees and to the Bayesian posterior distribution trees.
%The Maximum Likelihood bootstrap trees and the Bayesian posterior distribution trees are the most similar for both metrics.

@@

@@

@@

@@

% NC: Make sure you're consistent in how you refer to things ie Maximum Likelihood bootstrap trees, not Bootstraps

%---------------------------------------------
%
%       DISCUSSION
%
%---------------------------------------------
